Covariate analysis of CMC RNASeq data reprocessed with RSEM-feature counts workflow with GRCh38 (DLPFC with CQN normlaisation)

### 2018-10-25 code/Rmd/CMC_DLPFC_RSEM_GRCh38_CQN_ModSel.Rmd
sex-specific chromosome gene expression plots, MDS plot, Reported_Gender as a variable in gene filtering

knit2synapse::createAndKnitToFolderEntity(file = "./CMC_DLPFC_RSEM_GRCh38_CQN_ModSel.Rmd",
                                          parentId ="syn9846623",
                                          folderName = 'DLPFC - RSEM - EXPECTED COUNTS - GRCh38 - CQN - Sparse Model')
```

Notes: 
- 3 Kleinfelter individuals

@@

library(ggplot2)
library(limma)
library(psych)
library(edgeR)
library(cqn)
library(lme4)
library(biomaRt)
library(ComplexHeatmap)

@@

#cl = makeCluster(detectCores()-2)
cl = makeCluster(detectCores())
registerDoParallel(cl)

@@

# Github link
thisRepo <- getRepo(repository = "kelshmo/Brain_Reg_Net", ref="branch", refName='CMC')
thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/Rmd/',thisFileName))
```
### Data download
#### Obtain effective count matrix and metadata from synapse.

#readRDS(file = "logcpm.rds")
```{r download.data, cache=TRUE}
downloadFile <- function(id){
  fread(synGet(id)$path, data.table = F)
}
downloadFile_version <- function(id , version){
  fread(synGet(id, version = version)$path, data.table = F)
}
# Download reprocessed counts (DLPFC)
COUNT_ID = 'syn16783526'
ALL_USED_IDs = COUNT_ID
COUNT = downloadFile(COUNT_ID) %>% data.frame()
rownames(COUNT) = COUNT$gene_id
COUNT$gene_id = NULL
COUNT$transcript_id.s. = NULL

@@

GENE.GC.CONT = Ensemble2HGNC %>%
  dplyr::left_join(backgroundGenes) %>%
  dplyr::select(gene_id, percentage_gene_gc_content, chromosome_name) %>%
  unique
rownames(GENE.GC.CONT) = GENE.GC.CONT$gene_id
```

@@

```{r cpmnormalisation}
genesToAnalyze = dlply(METADATA, .(Dx), .fun = function(mtd, count){
  processed.counts = getGeneFilteredGeneExprMatrix(count[,mtd$SampleID],
                                                   MIN_GENE_CPM=1, 
                                                   MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0.5)
  processed.counts$filteredExprMatrix$genes
}, COUNT)

@@

```{r sex-specific_expr, fig.height=8, fig.width=8, results='asis', cache = FALSE}
#filter for sex specific chromosomes here. Plot x and y expression separately for males and females 
```

@@

### Library Normalisation
Initial library normalisation usign cqn
```{r cqn}
# Compute offset for gene length and gc content
GENE.LEN = tibble::column_to_rownames(GENE.LEN, var = 'gene_id') %>% as.data.frame()
CQN.GENE_EXPRESSION = cqn(NEW.COUNTS, 
                          x = GENE.GC.CONT[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'percentage_gene_gc_content'],
                          lengths = GENE.LEN[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'Length'],
                          lengthMethod = "smooth", 
                          verbose = FALSE)
CQN.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset

@@

### Sample clustering
```{r decompse.normalise.data, fig.height=8, fig.width=8, results='asis', cache=FALSE}
#plotMDS of CQN normalized expression data plotMDS(x) where x is any data object which can be coerced to a matrix

# Find principal components of expression to plot
tmp = CQN.GENE_EXPRESSION$E
tmp[is.na(tmp)] = 0
PC <- prcomp(tmp, scale.=T, center = T)

@@

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=Institution, shape=Dx, size=Age_of_Death))
p <- p + theme_bw() %+replace% theme(legend.position="top")
# p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
p
```
Tree based clustering of samples
```{r decompse.normalise.data.1, fig.height=6, fig.width=10, results='asis'}
# Eucledian tree based analysis
COVARIATES.tmp = data.matrix(COVARIATES[,c("Institution", "Gender", "LibraryBatch", "Dx")])
COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0

@@

# Download all related genesets from synapse
geneset.files = synQuery('select * from file where parentId == "syn3240583"')$results
genesets.to.test = purrr::map(geneset.files, .f = function(fileList){
  tmp = downloadFile(fileList$file.id) %>%
    dlply(.(Name), .fun = function(x){
      str_split(x$symBeforeOverlap, '\\|') %>% unlist %>% unique
    })
  }) %>%
  CovariateAnalysis::filterGeneSets(backgroundGenes) %>%
  do.call(c,.)

@@

# Code
CODE <- Folder(name = "DLPFC - RSEM - EXPECTED COUNTS - GRCh38 - CQN - Sparse Model", parentId = parentId)
CODE <- synStore(CODE)

>>>>

knit2synapse::createAndKnitToFolderEntity(file = "./CMC_DLPFC_RSEM_GRCh38_CQN_ModSel.Rmd",
                                          parentId ="syn9846623",
                                          folderName = 'DLPFC - RSEM - EXPECTED COUNTS - GRCh38 - CQN - Sparse Model - Kelsey')
```

@@

library(ggplot2)
library(limma)
library(psych)
library(edgeR)
library(cqn)
library(lme4)
library(biomaRt)
library(ComplexHeatmap)
library(statmod)

@@

cl = makeCluster(detectCores()-2)

registerDoParallel(cl)

@@

# Github link
thisRepo <- getRepo(repository = "kelshmo/Brain_Reg_Net", ref="branch", refName='CMC')
thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/Rmd/',thisFileName))
```
### Data download
#### Obtain effective count matrix and metadata from synapse.
```{r download.data, cache=TRUE}
downloadFile <- function(id){
  fread(synGet(id)$path, data.table = F)
}
downloadFile_version <- function(id , version){
  fread(synGet(id, version = version)$path, data.table = F)
}
# Download reprocessed counts (DLPFC)
COUNT_ID = 'syn16783526'
ALL_USED_IDs = COUNT_ID
COUNT = downloadFile(COUNT_ID) %>% data.frame()
rownames(COUNT) = COUNT$gene_id
COUNT$gene_id = NULL
COUNT$transcript_id.s. = NULL

@@

GENE.GC.CONT = Ensemble2HGNC %>%
  dplyr::left_join(backgroundGenes) %>% 
  dplyr::select(gene_id, percentage_gene_gc_content, chromosome_name) %>%
  unique
rownames(GENE.GC.CONT) = GENE.GC.CONT$gene_id
```
### Sex Chromosome-specific Gene Expression Patterns

```{r Reported_Gender_plots}
COUNT$gene_id = rownames(COUNT) 
REPORTED.GENDER.COUNTS = GENE.GC.CONT %>% 
  left_join(COUNT) %>%
  dplyr::select(-one_of("percentage_gene_gc_content")) %>%
  filter(chromosome_name == "X" |chromosome_name == "Y") %>% 
  tidyr::gather(key = item, value = value, -c(gene_id, chromosome_name)) %>%
  mutate(value = log(value)) %>%
  rename(`counts(log)`= value) %>% 
  rename(SampleID = item) %>%
  left_join(METADATA[,c("SampleID", "Reported_Gender")]) %>% 
  rename(`Reported Gender` = Reported_Gender)

my.theme <- theme_bw() %+replace% theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1), plot.title=element_text(hjust=0.5))

p = list()
p[[1]] = ggplot(filter(REPORTED.GENDER.COUNTS, chromosome_name == "X"), aes(x = `Reported Gender`, y = `counts(log)`)) + geom_boxplot()
p[[1]] = p[[1]] + ggtitle('X') + my.theme
p[[2]] = ggplot(filter(REPORTED.GENDER.COUNTS, chromosome_name == "Y"), aes(x = `Reported Gender`, y = `counts(log)`)) + geom_boxplot()
p[[2]] = p[[2]] + ggtitle('Y') + my.theme
multiplot(plotlist = p, cols = 2)
```

@@

```{r cpmnormalisation}
rownames(COUNT) = COUNT$gene_id
COUNT$gene_id = NULL

genesToAnalyze = dlply(METADATA, .(Dx, Reported_Gender), .fun = function(mtd, count){
  processed.counts = getGeneFilteredGeneExprMatrix(count[,mtd$SampleID],
                                                   MIN_GENE_CPM=1, 
                                                   MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0.5)
  processed.counts$filteredExprMatrix$genes
}, COUNT)

@@

@@

```{r testCPM}
# TEMP <- LOG.CPM
# TEMP$gene_id = rownames(TEMP) 
# t <- data.frame(TEMP) %>% 
#   mutate(gene_id = rownames(TEMP))
# REPORTED.GENDER.COUNTS = GENE.GC.CONT %>% 
#   left_join(t) %>%
#   dplyr::select(-one_of("percentage_gene_gc_content")) %>%
#   filter(chromosome_name == "X" |chromosome_name == "Y") %>% 
#   tidyr::gather(key = item, value = value, -c(gene_id, chromosome_name)) %>%
#   rename(`CPM(log)`= value) %>% 
#   rename(SampleID = item) %>%
#   left_join(METADATA[,c("SampleID", "Reported_Gender")]) %>% 
#   rename(`Reported Gender` = Reported_Gender)

# my.theme <- theme_bw() %+replace% theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1), plot.title=element_text(hjust=0.5))
# 
# p = list()
# p[[1]] = ggplot(filter(REPORTED.GENDER.COUNTS, chromosome_name == "X"), aes(x = `Reported Gender`, y = `CPM(log)`)) + geom_boxplot()
# p[[1]] = p[[1]] + ggtitle('X') + my.theme
# p[[2]] = ggplot(filter(REPORTED.GENDER.COUNTS, chromosome_name == "Y"), aes(x = `Reported Gender`, y = `CPM(log)`)) + geom_boxplot()
# p[[2]] = p[[2]] + ggtitle('Y') + my.theme
#multiplot(plotlist = p, cols = 2)
```
### Library Normalisation
Initial library normalisation usign cqn
```{r cqn}
# Compute offset for gene length and gc content
GENE.LEN = tibble::column_to_rownames(GENE.LEN, var = 'gene_id') %>% as.data.frame()
CQN.GENE_EXPRESSION = cqn(NEW.COUNTS, 
                          x = GENE.GC.CONT[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'percentage_gene_gc_content'],
                          lengths = GENE.LEN[PROCESSED_COUNTS$filteredExprMatrix$genes$genes, 'Length'],
                          lengthMethod = "smooth", 
                          verbose = FALSE)
CQN.GENE_EXPRESSION$E = CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset

@@

### Sample clustering
```{r decompse.normalise.data, fig.height=8, fig.width=8, results='asis', cache=FALSE}
# Find principal components of expression to plot
tmp = CQN.GENE_EXPRESSION$E
tmp[is.na(tmp)] = 0
PC <- prcomp(tmp, scale.=T, center = T)

p = list()
# Plot first 2 PCs
plotdata <- data.frame(SampleID=rownames(PC$rotation), 
                       PC1=PC$rotation[,1], 
                       PC2=PC$rotation[,2])

@@

p[[1]] <- ggplot(plotdata, aes(x=PC1, y=PC2))
p[[1]] <- p[[1]] + geom_point(aes(color=Institution, shape=Dx, size=Age_of_Death))
p[[1]] <- p[[1]] + theme_bw() %+replace% theme(legend.position="top") + guides(colour = guide_legend(nrow = 2),
                                                                               shape = guide_legend(nrow = 2),
                                                                               size = guide_legend(nrow = 2))
# p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)

#MDS
tmp.mds <- dist(t(tmp))
mds <- cmdscale(tmp.mds)
#Plot MDS
plotdata <- data.frame(SampleID = rownames(mds),
                       MDS1= mds[,1],
                       MDS2= mds[,2])

p[[2]] <- ggplot(plotdata, aes(MDS1,MDS2))
p[[2]] <- p[[2]] + geom_point(aes(color=Institution, shape=Dx, size=Age_of_Death))
p[[2]] <- p[[2]] + theme_bw() %+replace% theme(legend.position="top") + guides(colour = guide_legend(nrow = 2),
                                                                               shape = guide_legend(nrow = 2),
                                                                               size = guide_legend(nrow = 2))
multiplot(plotlist = p, cols = 1)
```
Tree based clustering of samples
```{r decompse.normalise.data.1, fig.height=6, fig.width=10, results='asis'}
# Eucledian tree based analysis
COVARIATES.tmp = data.matrix(COVARIATES[,c("Institution", "Reported_Gender", "LibraryBatch", "Dx")])
COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0

@@

# Download all related genesets from synapse
geneset.files = synGetChildren("syn3240583", includeTypes=list("file"))$asList()
genesets.to.test = purrr::map(geneset.files, .f = function(fileList){
  tmp = downloadFile(fileList$id) %>%
    dlply(.(Name), .fun = function(x){
      str_split(x$symBeforeOverlap, '\\|') %>% unlist %>% unique
    })
  }) %>%
  CovariateAnalysis::filterGeneSets(backgroundGenes) %>%
  do.call(c,.)

@@

# Code
CODE <- Folder(name = "DLPFC - RSEM - EXPECTED COUNTS - GRCh38 - CQN - Sparse Model - Kelsey", parentId = parentId)
CODE <- synStore(CODE)
